Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-02-23 iso-int/rc-iso-int.tex
this will be the rc2012 paper
fiore
_
outline

\begin{document}
\title{On the Construction of Isomorphic Interpreters} 
\titlerunning{On the construction of Isomorphic Interpreters} 
\author{Roshan P. James \and Amr Sabry}
\institute{School of Informatics and Computing, Indiana University\\
\mails}
\maketitle

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Logically Reversible Multiplier}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A {{langRev}} Interpreter}

%subcode{bnf} include main
% Combinators, c = iso | c (;) c | c (*) c | c (+) c
% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
%                  &|& LeftTimes cc c v | RightTimes c v cc 
%                  &|& LeftPlus cc c | RightPlus c cc 
% Values, v = () | (v, v) | L v | R v
%
% Machine states = <c, v, cc> | {[c, v, cc]}
% Start state = <c, v, []> 
% Stop State = {[c, v, []]}

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi1.pdf}
}
\end{center}

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi2.pdf}
}
\end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A {{langRevEE}} interpreter}

%subcode{bnf} include main
% Combinators, c = iso | eta | eps | c (;) c | c (*) c | c (+) c
% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
%                  &|& LeftTimes cc c v | RightTimes c v cc 
%                  &|& LeftPlus cc c | RightPlus c cc 
% Values, v = () | (v, v) | L v | R v
%
% Machine states = <c, v, cc> | {[c, v, cc]} | <c, v, cc>^{dagger} | {[c, v, cc]}^{dagger}
% Start state = <c, v, []> 
% Stop State = {[c, v, []]}
%

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
% <eps, R v, cc> &|-->& <eps, L (-v), cc>^{dagger} &~~~~~~~~~~~~ rule 12
% <eps, L (-v), cc> &|-->& <eps, R v, cc>^{dagger} &~~~~~~~~~~~~ rule 13

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc>^{dagger} &<--|& {[iso, iso(v), cc]}^{dagger} &~~~~~~~~~~~~ rule 1'
% <c1(;)c2, v, cc>^{dagger} &<--|& <c1, v, Fst cc c2>^{dagger} &~~~~~~~~~~~~ rule 2'
% {[c1, v, Fst cc c2]}^{dagger} &<--|& <c2, v, Snd c1 cc>^{dagger} &~~~~~~~~~~~~ rule 3'
% {[c2, v, Snd c1 cc]}^{dagger} &<--|& {[ c1(;)c2, v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 4'
% <c1(+)c2, L v, cc>^{dagger} &<--|& <c1, v, LeftPlus cc c2>^{dagger} &~~~~~~~~~~~~ rule 5'
% {[ c1, v, LeftPlus cc c2 ]}^{dagger} &<--|& {[c1 (+) c2, L v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 6'
% <c1(+)c2, R v, cc>^{dagger} &<--|& <c2, v, RightPlus c1 cc>^{dagger} &~~~~~~~~~~~~ rule 7'
% {[ c2, v, RightPlus c1 cc ]}^{dagger} &<--|& {[c1 (+) c2, R v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 8'
% <c1(*)c2, (v1, v2), cc>^{dagger} &<--|& <c1, v1, LeftTimes cc c2 v2>^{dagger} &~~~~~~~~~~~~ rule 9'
% {[ c1, v1, LeftTimes cc c2 v2 ]}^{dagger} &<--|& <c2, v2, RightTimes c1 v1 cc>^{dagger} &~~~~~~~~~~~~ rule 10'
% {[ c2, v2, RightTimes c1 v1 cc ]}^{dagger} &<--|& {[ c1 (*) c2, (v1, v2), cc ]}^{dagger} &~~~~~~~~~~~~ rule 11'
% {[eps, L (-v), cc]}  &<--|& {[eps, R v, cc]}^{dagger}  &~~~~~~~~~~~~ rule 12'
% {[eps, R v, cc]} &<--|&  {[eps, L (-v), cc]}^{dagger} &~~~~~~~~~~~~ rule 13'

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/piee1.pdf}
}
\end{center}

Let us summarize it as follows.

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/piee2.pdf}
}
\end{center}

Adding backward execution using the adjoint.

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/piee3.pdf}
}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{CEK machine}

%% %subcode{bnf} include main
%% % e = x | \x.e | e e
%% % rho = [] | x=v:rho
%% % E = [] | (rho, e):E | v:E
%% % v = (\x.e, rho)
%% %
%% % Machine states = <e, rho, E>
%% % Start state = <e, [], []>
%% % Stop state = <v, rho, []>

%% %subcode{opsem} include main
%% % <e1 e2, rho, E > &|-->& <e1, rho, (rho, e2):E >
%% % <\x.e, rho, (rho', e'):E > &|-->& <e', rho', (\x.e,rho):E >
%% % <\x.e, rho, (\x'.e',rho'):E > &|-->& <e', x' =(\x.e,rho):rho', E >

>>>>

\begin{document}
\title{Isomorphic Interpreters from Logically Reversible Abstract Machines} 
\titlerunning{On the construction of Isomorphic Interpreters} 
\author{Roshan P. James \and Amr Sabry}
\institute{School of Informatics and Computing, Indiana University\\
\mails}
\maketitle

@@

Main points: 
\begin{itemize}
\item Make the case for reversibility via type isomorphisms
\item Type isomorphisms between finite types are straightforward (but
  tedious)
\item Type isomorphisms between recursive types are quite difficult. This is
  what we focus on.
\end{itemize}
To understand the problem in more detail, consider the problem of converting
a reversible recursive function into a partial isomorphism. In general such a
function takes an input, makes several recursive calls, and produces an
answer. To implement this in a reversible way, the structure of the recursion
and the number of recursive calls need to be exposed. The key trick is to
express tease the function into two components: an outer driver that performs
iteration and an inner state transition function that realizes one
``unfolding'' of the recursion. In program transformation world, this is
called, light-weight fission. Using this idea, one can realize any recursive
function as a partial isomorphism. We show a few examples of the construction
culminating with the most challenging recursive function: a full interpreter.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A {{langRevT}} Interpreter}

%subcode{bnf} include main
% Combinators, c = iso | c (;) c | c (*) c | c (+) c | trace c
% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
%                  &|& LeftTimes cc c v | RightTimes c v cc 
%                  &|& LeftPlus cc c | RightPlus c cc | Trace cc
% Values, v = () | (v, v) | L v | R v
%
% Machine states = <c, v, cc> | {[c, v, cc]}
% Start state = <c, v, []> 
% Stop State = {[c, v, []]}

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
% < trace c, v, cc> &|-->& <c, R v, Trace cc> &~~~~~~~~~~~~ rule 12
% {[c, L v, Trace cc]} &|-->& <c, L v, Trace cc> &~~~~~~~~~~~~ rule 13
% {[c, R v, Trace cc]} &|-->& {[trace c, R v, cc]} &~~~~~~~~~~~~ rule 14

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi0-1.pdf}
}
\end{center}

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/pi0-2.pdf}
}
\end{center}

@@
